---
title: "R Introduction"
output: html_notebook
---


```{r setup}
# check required pacakges are installed
cran_inst <- setdiff(c('tidyverse', 'fortunes', 'dslabs'), rownames(installed.packages()))
if (length(cran_inst) > 0) { install.packages(cran_inst) }
# load libraries
library(tidyverse)
```

### R Notebooks
* Such as this one, combine markdown text with R code
* A single line of R code can be run by place the cursor on the line and pressing `ctrl+enter`/`cmd+enter`
* Code can also be entered directly into the R console, and run by pressing enter
* R code "chunks", such as the one below, can be run by:
  1. clicking the `run` button at the right of the chunk
  2. placing the cursor inside the chunk then pressing `ctrl+shift+enter`/`cmd+shift+enter`


```{r}
# random quotes from R developers
fortunes::fortune()
```
* A new code chunk may be inserted by
  1. pressing `ctrl+alt+i`/`cmd+opt+i`
  2. clicking `Code->Insert Chunk` in the top menu

### Maths
* R can be used to perform basic mathematic operations

```{r}

```

* Brackets are used to control to control the order of operations

```{r}

```

### Variables
* data is assigned to variables with the `<-` assignment operator
* varibles may be directly assigned with a value, or with the result of an expression
* variables can then be used in subsequent expressions

```{r}

```

* in addition to numbers, variables may store character "strings" (a sequence of characters) or logical values (TRUE or FALSE)

```{r }

```

### Variable names
* variable names may be made up of upper and lower case letters, numbers and underscores
* variable names must start with a letter
* good variable names are short and descriptive
* use underscores in place of spaces to separate words in variables

```{r}

```

### Vectors
* Many variables R are vectors, and may contain multiple values of the same type
* We can access individual vector elements by their index
  * e.g. for the first element of vector `x` we use `x[1]`
* the `length(x)` function will return the length of a vector
* When we use a vector in a mathematical expression, the formula is applied to each element of the vector

```{r variables_3}

```

* two or more vectors may be combined with the `c()` function

```{r}

```


* we can call summary functions on vectors to obtain a single value
* examples of summary functions are `min(x)`, `max(x)`, `mean(x)`, `sd(x)`

```{r}

```

#### Missing values
* missing values are indicated by `NA` in R

```{r}

```

#### Exercise
1. Calculate the mean of z, excluding NA values. Refer to the documentation of mean by typing `help(mean)` into the console.
```{r}

```


### Type conversions
* the function `class()` will return the type of a given vector
* vectors may be converted from one type to another with `as` functions
* Examples include `as.integer()`, `as.numeric()`, `as.character()`, `as.logical()`
* When we use `c()` to combine different vectors all elements are converted to the same type

#### Exercise
1. use the `class()` function to determine the type of the following variables
2. use the `c()` function to combine the variables into all pairwise combinations. 
Determine the classes of the resulting vector, and establish the rules for type conversion.
```{r}
foo <- 1.5
bar <- 5L
baz <- 'apple'
zum <- TRUE


```

### Lists
* Lists in R can stored named collections of objects of different types
* Elements of a list can be accessed by name with the `$` operator
```{r}

```

### Data frames
* tabular data in R is stored in data frames, which are similar to spreadsheets
* Data frames are a special kind of list that store vectors of the same length, which each vector making up a column in the data frame
* a `tibble` is a type of data frame that is cleaner to work with
* We can view a data frame with the Rstudio viewer by clicking the name of the data fram in the top right environment tab
* We can access specific columns with the `$`, as with lists
* We can get the number or rows and the number of columns from a data frame with `nrow(x)` and `ncol(x)`

#### Exercise
* use `nrow`, `ncol` and `$` on the data frame `mpg_head` below
```{r dataframes}
mpg_head <- head(ggplot2::mpg)
```

### Plotting Data
* We can plot data in a dataframe using ggplot
* We do this by adding geoms (geometric representations of data) to a base plot
* The data to be plotted is specified using `aes()`
* see the ggplot2 cheatsheat for common geoms

```{r}
xy_data <- tibble(x = seq.int(-10, 10), y= x ** 2)

ggplot(xy_data) + 
  geom_line(aes(x = x, y = y)) 

ggplot(xy_data) + 
  geom_line(aes(x = x, y = y)) + 
  geom_point(aes(x = x, y = y))
```


#### Excercise
1. Fill in the missing arguments in `aes` to give a scatter plot where the x axis show `displ` and the y axis shows `hwy`.
  - see `help(geom_point)`, the examples at the bottom of the help page are most useful
2. Add an argument to `aes()` to colour the points by number of cylinders
3. Add descriptive axis labels and a title to the plot with `labs()`


```{r}
# note: the mpg data frame contains data of on the miles per gallon of various makes of car
# hwy is the highway mileage, displ is the engine displacement (size) and cyl is the number of cylinders in the engine
mpg <- ggplot2::mpg

ggplot(mpg) + geom_point(aes( ))
```

### Manipulating data frames
* The `select` function is used to select a set of columns in a data frame
* The `filter` function subsets a data frame for rows where the supplied conditions is `TRUE`
* The pipe `%>%` function can be used to chain several function calls into a pipeline

```{r}

# select columns from a dataframe 
select(mpg, cyl, displ, hwy)

# the pipe - takes the previous result and use it in the next function
mpg %>% select(cyl, displ, hwy)
# start with mpg then select cyl, cty and hwy

# filter for rows that match a condition
mpg %>% filter(cyl < 8)
mpg %>% filter(cyl < 8, manufacturer != 'audi')
mpg %>% filter(cyl < 8, manufacturer != 'audi', year == 2008)

```

#### Exercise
1. Combine the `select`, `filter` and `ggplot` functions into a pipeline with the `%>%` operator with mpg dataset
```{r}

```


* `mutate` allows us to add new columns to a data frame, often as some function of existing columns

```{r}
xyz_data <- 
  tibble(x = seq.int(-10, 10), y = x ** 2) %>% 
  mutate(z = x + y)

xyz_data
```


#### Exercise
1. add a column named "rate" to the contagious data frame that divides the count by the population using `mutate`
2. `filter` the resulting data frame for observations in the year 1975
3. make a plot using `geom_boxplot` showing the distribution of rate by disease
4. add descriptive titles with `labs`
```{r transform_data}
# data on counts of reported infectious disease in US states b year
contagious <- as_tibble(dslabs::us_contagious_diseases)

```

### Summarising data
* `summarise()` computes summary statistics from a dataframe
* This is used together with `group_by()` to define which subsets of the data frame to summarise over
* A summary function used in summarise must return a single value for each group
* All columns not used for grouping will be dropped from the resulting data frame

```{r}
mpg %>% 
  group_by(cyl) %>% 
  summarise(mean_cty = mean(cty), mean_hwy = mean(hwy))
```

* `count()` operates similarly to to summarise, and returns the number of members of a grouping given `group_by`

#### Exercise 
1. use `group_by` and `count()` to count the number of models by "manufacturer" and "cyl" from mpg

```{r}

```



#### Exercise
1. use `group_by` and `summarise` on the contagious data frame to calculate the mean disease rate by disease and year
2. plot the resulting summary with `geom_line`

```{r}

```


### Exercise
Explore the gapminder data set
1. Add a column "gpd_per_cap" by dividing "gdp" by "population"
2. Summarise means for each of the variables by year and region
  a. create a plot using `geom_col` showing one of these variables for the year 1960
  b. use `facet_wrap` to plot this variable at years 1960, 1970, 1980, 1990 and 2000.
  Hint: use `filter(year %in% c(1960, 1970, 1980, 1990, 2000))` to select the correct years
3. create a histogram with `geom_histogram` of "gpd_per_cap" in the year 2007
4. create a line plot for fertility rates in the Americas colored by country
5. create a plot showing "gdp_per_cap" on the x axis and  "life_expectancy" on the y axis in the year 2007
  a. add aesthetics to this plot that show 1) continent and 2) population

```{r}
gapminder <- as_tibble(dslabs::gapminder)

```

